### 2025-01-05 reviewer-response.md

Dear Dr. Becker,

We thank you and the anonymous reviewers for the thoughtful and insightful suggestions regarding our manuscript. Below, please find a point-by-point accounting of our responses to the reviews.

Associate Editor comments (Remarks to Author):
I have sent the manuscript out to two referees and received those reports back, and I have also looked at your paper myself. Both referees have identified quite a few items they would like to see addressed, but overall I'm pleased to report that I think the manuscript has merit and impact, and so I am recommending a major revision. In addition to rephrasing, reorganizing and clarifying the paper (per the referees' comments), the two major items to address are (1) details about the real world examples (cf comments from referee 1), and (2) referee 2 raises an issue about different ground truths.

**Referee #2 (Remarks to the Author):**

*In "Let them have CAKES [...]", the authors describe a new method to perform
variants of the nearest neighbor search. The Rust implementation of this new
algorithm is relatively (comparable or somewhat slower than the competition) but
has a much better recall across many different tasks.*

*The description of the method is clear, although the description of the run time
costs is ambiguous. Overall, CAKES is an interesting contribution to the field
of NN-search.*

@@

# Major comments

*The big O notation by default suggest worst-case run time complexity. In this
manuscript, it seems that complexity is a mixture of expected run time
complexity and under the loose assumption that the data has particular
characteristics (e.g., low fractal dimension, the tree is balanced). These
assumptions should be made clear in the intro and when giving the run time
expressions.*

this is easy to address, and we will do so

*For example, line 215, does the cost of partitioning of O(n log(n)) depend on
the tree being balanced? It is an important hypothesis because line 197 states
that "the imbalance in the tree is a feature, not a bug, as it reflects the
underlying structure of the data". That may look like a contradiction.*

we can address this with the plots that include a balanced tree -- expand 2.1.3, balanced vs unbalanced, dropping off points faster with an unbalanced tree?

*The cost on line 260 uses \hat{r}, which is a mean radius. Now that cost is in
expectation (in addition to the balanced tree hypothesis).*

this is restating the complexity from the CHESS paper, but it's correct that it's an expected cost. Worst-case analysis isn't really useful here.

*Another example, line 326, comes from the minimum with value 2 on line 284. If
the assumption on the LFD is not satisfied, then an extra log factor would
appear in the cost.*

write in plain english right after eq 2.3, explaining in plain english the rationale for the min(2,x) expansion: namely, we don't think the very low LFD would be representative of what we'd see if we massively increased the radius

*The discussion on LFD after its definition line 141 is hard to understand:
there are 2 parameters in the definition of LFD (r_1 and r_2), but the
discussion seems to imply there is only 1 (a radius r). That discussion only
seems to make sense after line 163 and the rewrite of the definition 2.1.*

Right after or in eq 2.1, immediately follow up with the simplified form where $r_1 = 2r_2$. Also drop subscript $X$, since we're dealing with only one dataset. Explain.

*Smith-Waterman and Needleman-Wunsch are not distances (unlike Levenshtein and
Hamming distances). They are dynamic programming algorithms to compute
distances like the Levenshtein distance (and others depending on the exact
parameters of the algorithms).*

Correct; fix in language.

*The explanation that the augmentation process (starting at line 418) preserves
the topological structure of the underlying data is not very convincing. This
process adds small clouds of points extending in every dimension around
existing points. That is not quite the same as generating new points that
belong to a small dimension manifold in the sample space.*

We should acknowledge this limitation, and suggest that in future work we would do something like favor augmentation along the first few PCs of the local manifold, but not in scope here. This goes in discussion.

*Line 467-473: Why are there multiple linear search implementations used as
ground truth? These different implementation are apparently used to evaluate
different algorithms (ANNOY, FAISS-IVF vs CAKES). This is troubling: why would
the ground truth differ based on which tool is evaluated or which language it
is written in?*

*Do these 2 tools return the same result? If so, why use multiple tools? If
not, then the evaluation is flawed.*

We don't have a python wrapper for cakes, and the other tools were in python. There is only one ground truth. Fix language starting in Results.

@@

# Minor comments

*Line 421: The points are created with distance $\epsilon$ of $x$, not $\epsilon
||x||$. $r$ is chosen in the sphere of radius $\epsilon$ and added to $x$, and, as
stated correctly line 424, $x'$ is within $\epsilon$ of $x$.*

x is a vector, and epsilon is a scalar. We can clarify that x is a vector. Be sure to boldface all vectors

*There are many small LaTeX issues: wrong space after "i.e." and "e.g." (add a
backslash), large spaces around formulas (e.g., line 342), em-dashes that look
too short (e.g., line 594).*

easy to fix (NMD)

*The figures should not be shrank so much: the labels on the axis and the keys
are not readable. Regenerate the pictures at the correct size to avoid
scaling.*

fix some font sizes and aspect ratios

*Put the keys outside of the figures, not over the data lines.*

need to fix this (NI)

*Many horizontal and vertical lines in Table~2 should be removed to help with
readability.*

investigate this, or break up table 2 into 2a-2d (NMD)

*Line 585: "we observe performs quite slowly". The sentence seems incomplete.*

will fix -> 'we observe that CAKES performes quite slowly'

*Many of the explicit algorithms do not serve any real purpose: they are not
more readable in pseudo-code form and are just fine in explained in the text.
(E.g., Algorithm 2.4).*

get rid of alg 2.4

*The Github repository gives very little instructions on how to compile the
code, run it or reproduce the results from this manuscript.*

will fix (NI)

**Review of
Let them have CAKES: A Cutting-Edge Algorithm for Scalable,
Efficient, and Exact Search on Big Data**

*SIMODS manuscript M166272*

@@

MAJOR ISSUES =============================================================

*The CAKES manuscript proposes an entropy scaling approach to the search
problems of $k$ Nearest Neighbors ($k$-NN) and $rho$ Nearest Neighbors
($\rho-NN) in a data set $X$.  The general approach is to construct a
hierarchical data structure of clusters to assist in the searching.
While there are interesting ideas in the manuscript, there are major
issues both technically and expositionally, as well as some minor issues,
that need to be addressed before another round of reviewing.*

*While there is implicitly a tree that is the key data structure, the tree
itself is never named or given a notation; the same is true of the root of
the tree.  The authors should review the requirements for defining proper
terminology and notation for a data structure and apply it consistently to
their data structure.*

we fixed this in pancakes, just backport it

*The authors appear to use the terms similarity and distance
interchangeably, though they are certainly different.  In particular, a
similarity is a measure (nominally, non-negative) that is (ideally) larger
the closer two data points are and, in some biological contexts, has no
upper bound.  Moreover, the use of a distance measure alone is inadequate
for the described algorithms.  On line 155, there is mention of the
geometric median of the points, but there may be no such median depending
on the space (set) containing the points of the data set.  For example,
there is no obvious median for a set of biological sequences.  The authors
must identify the actual spaces from which data sets can be extracted for
CAKES to work.*

add def in paper for geometric median, extended from continues vector spaces into any metric space. line 43, use distance, not similarity

*On line 141, there is a function defined for the local fractal dimension
(LFD), but it is not expressed in functional notation.  In other words, it
should start with $LFD(X,r_1,r_2)$, since it is clearly a function of all
three.  This functional notation should propagate throughout the rest of
the manuscript.  Much the same is true on line 142, where it should read
$B(X,q,r)$, not $B_X(q,r)$.  If makes little sense to put $X$ as a
subscript in this context.*

similar to complaint from reviewer 2, address with same language

*In the algorithms, starting with Algorithm 2.1, the pseudocode style is
ugly.  Much better would be the pseudocode style from the fourth edition
of Introduction to Algorithms, CLRS.  Also, the description of $criteria$
is as a "stopping criteria", while the logic of the code has it as a
continuing criteria.  Please fix the pseudocode throughout.*

make this pseudocode style change (NMD)
call it continuation criteria (NI)

*Section 2.2 is meant to be the definition of the search problem(s), but
that is really only contained in the paragraph from lines 218 to 223. In
any case, it is essential to put the problem(s) to be solved earlier, that
is, first, before any description of the solutions with ideas such as
clusters.  This movement of text will require a bit of text refactoring at
a minimum.  Also, the sentence at lines 221 to 223 should be expanded in
full detail into a separate paragraph.  Finally, one assumes that the
integer $k$ or the number $\rho$ is a parameter to the corresponding
problem and should be mentioned as such.*

move introduction of the three deltas to later, to start of 2.4.3, and expand definition of rnn. (NI)

*There are no theorems in the manuscript, yet there are arguments for
quantities like the quality of the results or the time complexity that are
given without much precision.  It is better to give precise theorems or
propositions for each thing to be proved, with appropriate assumptions,
and to give precise proofs of same.*

state theorem, which is the time complexity, and formalize the argument into a proof

*The manuscript makes too much use of unfamiliar terminology taken from
earlier papers without defining terms.  For example, see line 233 and
Reference [18].  See also the paragraph at lines 306 to 310. All terms
used must be precisely defined in the manuscript to allow proper
assessment of the manuscript.*

remind the reader what rnn search is, and in 306-310 reword to refer to previous section rather than external refs

*The Results section is a slog to read and parts of the Discussion seem
redundant.  Take lines 651 to 668 as an example.  Tightening up the
Results and the Discussion can only help the exposition.*

offload to figure captions where possible, don't spell out everything

*I led a bioinformatics project that used an entropy scaling approach to
search a database of protein sequences.  See Reference [1].  As a result,
I am confused about some of the real world examples.  I know that protein
sequences are challenging to search for reasons such as their having
widely varying lengths and there being no obvious notion of a center of a
set of protein sequences.  In any case, I would like the authors to
reference [1] and discuss in more depth how they address the challenges of
the various real world examples.*

fine, cite 1, and discuss (NI)

@@

MINOR ISSUES =============================================================

*"Data set" is pronounced as two words and should be written as such
throughout the manuscript.*

We respectfully disagree; this appears as a term of art in existing literature, and is in the Merriam-Webster Dictionary. "Dataset" is listed by Merriam-Webster as the primary form, with "data set" being the less common form (https://www.merriam-webster.com/dictionary/dataset). Unless this is an editorial policy, we would rather keep it as "dataset."

*In line 22, use "rate" instead of "scale".*

ok, will fix

*Throughout, the word "which" is used when "that" is the correct word.  If
"which" is used, then it is not defining and is preceded by a comma.
Change all "which"'s to "that"'s.*

We somewhat agree, but disagree that *all* instances should be changed. We have made selected changes where it seems to fit the pattern suggested by this reviewer.

*On line 75, there is a Subsection 1.1, but there is no Subsection 1.2.
This is not a good practice.  Just refactor the text without Section 1.1
or add a Subsection 1.2.*

make the last paragraph (entropy scaling) into 1.2

*The use of $\forall$, $\exists$, $\Leftrightarrow$, and other similar
logical operators is not appropriate in a manuscript that is not about
logic.  Spell out the meaning in English, and be creative.*

We respectfully disagree. These operators are also standard in set theory and functional analysis, and it is in this context that we use them. They are also standard in the algorithms literature, including CLRS. Unless this is editorial policy, we would prefer to keep them. We believe they lend precision.

*Near line 154, it is important to give an overview of your solution
strategy and, in particular, why clustering is used first.  Clustering is
a heavily overloaded word.  I would have chosen a more neutral term
instead of "cluster".*

We do define our clustering approach, and moreover, we build on prior work (some ours, some others) that relies on clustering. We respectfully would prefer to stay with the language we have here.

*On line 155, there is the parenthetical comment about "a smaller sample of
the points".  What does that mean precisely and where does it occur later
in the manuscript?*

we take a root-n sample of the points (NI), justified in CHESS paper. Make this clear.

*On line 156, there is the claim "so it is a real data point".  What is a
real data point?  Please explain.*

deal with by defining geometric median

*On line 161 and elsewhere, "i.e." is always preceded and followed by
commas.*

fixed here

*On line 178, there is the phrase "find the point ... in the sample".  Can
you make the entire paragraph more precise by giving notation for the
sample?*

deal with by defining geometric median

*Section 2.1.1 overall is lacking in mathematical rigor.*

we will refactor 2.1.1 to be more rigorous (NI)

*In Section 2.1.3, the cost of assessing "criteria' is not counted in the
time complexity, but it must be.*

add some explanation: these can be as complex as a user wants, but for us it is a constant-time check (NI)

*In line 252, there is again mention of "real points" without any
indication of the meaning of "real".*

replace with 'actual points from the dataset'

*The fonts in Algorithms 2.2 through 2.7 are simply too small.  Make them
the font size of the remainder of the text.*

we will make this more legible (NMD)

*The numbers in Table 2 are too small.  Refactor into multiple tables
perhaps.*

we are refactoring into multiple tables (NMD)

*In line 641, what is "radial increase 2.3"?*

missing word 'equation' and also put in parentheses

*In line 642, why is "harmonic" italicized?*

no need

*The References section needs a through going over for proper formatting
and capitalization.  For example, check the title capitalization in [2],
[4], and [8].  Check the journal name capitalization in [9] and [21],
among others.*

*Several references have repeated URLs.  Trim these up.*

*References [2] and [3] appear to be duplicates.*

*Reference [39] is incomplete and inadequate.*

Will go over this carefully (NMD)

@@

REFERENCE ================================================================

*[1] Yoonjin Kim, Zhen Guo, Jeffrey A. Robertson, Benjamin Reidys, Ziyan
Zhang, Lenwood S. Heath, EnTrance: Exploration of Entropy Scaling Ball
Cover Search in Protein Sequences, bioRxiv
https://urldefense.com/v3/__https://www.biorxiv.org/content/10.1101/2021.05.31.446458v1__;!!Jh1S!kGffVkV6p8O_T-q3TzzRIBjJpAmGsw1kVTB7zosA7khu71pJpSxxrEqcYu6KceYMTg_VZ_MJ8B-PoPqf$ , 2021.*

### 2025-01-05 sections/2-methods.tex
dealt with e.g., i.e., dashes

We define a \textit{distance function} $f : \textbf{X} \times \textbf{X} \mapsto \mathbb{R}^+ \ \cup \ \{0\}$ which, given two points, deterministically returns a finite and non-negative real number.
A distance of zero defines identity among points (i.e. $f(x, y) = 0 \Leftrightarrow x = y$) and larger values indicate greater dissimilarity among points.
We also require that the distance function be symmetric (i.e., $f(x, y) = f(y, x) \ \forall \ x, y \in \textbf{X}$).
In addition to these constraints, if the distance function obeys the triangle inequality (i.e. $f(x, y) \leq f(x, z) + f(z, y) \ \forall \ x, y, z \in \textbf{X}$), then it is also a \textit{distance metric}.
Similar to~\cite{yu2015entropy}, when used with distance metrics, all search algorithms in CAKES are exact.
For example, Euclidean, Levenshtein~\cite{levenshtein1966binary} and Dynamic Time Warping (DTW)~\cite{muller2007dynamic} distances are all distance metrics, while cosine distance is not a metric because it violates the triangle inequality (e.g., consider the points $x = (1, 0)$, $y = (0, 1)$ and $z = (1, 1)$ on the Cartesian plane).

@@

Intuitively, LFD measures the rate of change in the number of points in a ball of radius $r$ around a point $q$ as $r$ increases. When the vast majority of points in the dataset have low ($\ll D$) LFD, we can simply say that the dataset has low LFD.
We stress that this concept differs from the \textit{embedding dimension} of a dataset.
To illustrate the difference, consider the SILVA 18S rRNA dataset which contains genomic sequences with unaligned lengths of up to 3,718 base pairs and aligned length of 50,000 base pairs.
Hence, the \textit{embedding dimension} of this dataset is at least 3,718 and at most 50,000.
However, physical constraints (namely, biological evolution and biochemistry) constrain the data to a lower-dimensional manifold within this space.
LFD is an approximation of the dimensionality of that lower-dimensional manifold in the ``vicinity'' of a given point.
Figure~\ref{fig:results:lfd-plots} illustrates this concept on a variety of datasets, showing how real datasets uphold the manifold hypothesis.
For real-world datasets, we expect the LFD to be locally uniform, i.e., when $r$ is small, but potentially highly variable at larger length scales, i.e., when $r$ is large.

@@

We define a \textit{cluster} as a set of points with a \textit{center} and a \textit{radius}.
The center is the geometric median of the points in the cluster, i.e., it is the point which minimizes the sum of distances to all other points in the cluster.
In cases where the cardinality of the cluster is large, we take a random subsample of $\sqrt{|C|}$ points and compute the geometric median of that subsample~\cite{ishaq2019clustered}.
The center, therefore, is one of the points in the cluster and is used as a representative of the cluster.
The radius is the maximum distance from the center to any point in the cluster.
Each non-leaf cluster has two child clusters in much the same way that a node in a binary tree has two child nodes.
Note that clusters can have overlapping volumes and, in such cases, points in the overlapping volume are assigned to exactly one of the overlapping clusters.
As a consequence, a cluster can be a proper subset of the metric ball at the same center and radius, i.e. $C(c, r) \subset B(c, r)$.

Hereafter, when we refer to the LFD of a cluster, it is estimated at the length scale of the cluster radius and half that radius, i.e. using Equation~\ref{eq:methods:lfd-half}.
We also only use points that are in $C(c, r)$ instead of all those in $B(c, r)$.

@@

The \textit{radius} of $C$ is the maximum distance from the \textit{center} to any other point in $C$.
The point which is responsible for that radius (i.e., the furthest point from the center) is designated the \textit{left pole} and the point which is furthest from the left pole is designated the \textit{right pole}.
We then partition the cluster into a \textit{left child} and a \textit{right child}, where the left child contains all points in the cluster which are closer to the left pole than to the right pole, and the right child contains all points in the cluster which are closer to the right pole than to the left pole.
Without loss of generality, we assign to the left child those points which are equidistant from the two poles.
Starting from a root cluster containing the entire dataset, we repeat this procedure until each leaf contains only one datum, or we meet some other user-specified stopping criteria, for example, minimum cluster radius, minimum cluster cardinality, maximum tree depth, etc.
This process is described in Algorithm \ref{alg:methods:partition}.
During the partitioning process, we also compute (and cache) the LFD of each cluster using Equation~\ref{eq:methods:lfd-half}.

@@

Note that Algorithm~\ref{alg:methods:partition} does not necessarily produce a balanced tree.
In fact, for real datasets, we expect anything but a balanced tree; the varying sampling density in different regions of the manifold and the low dimensional ``shape'' of the manifold itself will cause it to be unbalanced.
The only case in which we would expect a balanced tree is if the dataset were uniformly distributed, e.g. in a $d$-dimensional hyper-cube.
In this sense, the imbalance in the tree is a feature, not a bug, as it reflects the underlying structure of the data.

@@

This complexity analysis ignores the cost of checking the continuation criteria.
While a user could specify criteria of any complexity (our implementation allows for this), for this paper we continue clustering until each cluster is a singleton, i.e., it contains only one point or duplicates of the same point and has a radius of zero.
This criterion costs $\mathcal{O}(1)$ to check per cluster, and so does it not affect the overall complexity of the algorithm as used in this paper.

@@

Based on the assumption that the LFD near the query does not differ significantly from that of the clusters near the query (the set $Q$), Equation~\ref{eq:methods:repeated-rnn-factor} suggests that in the expected case, we need only two iterations of tree-search to find $k$ neighbors:
one iteration to find at least one cluster, and one more to find enough of the $k$ neighbors.
Since this is a constant factor, complexity of tree-search for Repeated $\rho$-NN is the same as that of $\rho$-NN search, i.e. $\mathcal{O}\big(\log\mathcal{N}_{\hat{r}}(X)\big)$.

@@

In addition, by our assumption that the LFD at the query is not significantly different from the LFD of nearby clusters, we have that $\hat{d} \approx d$.
By combining the bounds for tree-search and leaf-search, we see that Repeated $\rho$-NN has an asymptotic complexity of
\begin{gather}
    \mathcal{O}
    \Bigg(
        \underbrace{
            \log~\overbrace{\mathcal{N}_{\hat{r}}(X)}^{\textrm{metric entropy}}
        }_{\textrm{tree-search}}
        \ + \
        \underbrace{
            \overbrace{k}^{\textrm{output size}}
            \overbrace{\bigg( 1 + 2 \cdot \Big( \frac{\hat{|C|}}{k} \Big) ^ {d^{-1}} \bigg)^d}^{\textrm{scaling factor}}
        }_{\textrm{leaf-search}}
    \Bigg)
    \label{eq:methods:repeated-rnn-complexity}
\end{gather}
where $\mathcal{N}_{\hat{r}}(X)$ is the metric entropy of the dataset, $d$ is the LFD of the dataset, and $k$ is the number of nearest neighbors.
We note that the scaling factor should be close to 1 unless fractal dimension is highly variable in the region around the query (i.e. if $\hat{d}$ differs significantly from $d$).

@@

This process is described in Algorithm~\ref{alg:methods:depth-first-sieve}.
It terminates when $H$ is full and the top priority element in $H$ has $\delta$ less than the top priority element in $Q$, i.e., the theoretically closest point left to be considered in $Q$ is farther from the query than the $k^{th}$ nearest neighbor in $H$.
This leaves $H$ containing exactly the $k$ nearest neighbors to the query.

@@

We start with the assumption that the LFD near the query does not differ significantly from that of nearby clusters, and we consider leaf clusters with cardinalities near $k$.
Let $d$ be the LFD in this region.
Then the number of leaf-clusters in $Q$ is bounded above by $2d$, e.g. we could have a cluster overlapping the query ball at each end of each of $\lceil d \rceil$ mutually-orthogonal axes.
In the worst-case scenario for tree-search, these leaf clusters would all come from different branches of the tree, and so tree-search looks at $2 \cdot \lceil d \rceil \cdot \log \mathcal{N}_{\hat{r}}(X)$ clusters.
Thus, the asymptotic complexity is $\mathcal{T} \coloneqq \mathcal{O} \big( \lceil d \rceil \cdot \log \mathcal{N}_{\hat{r}}(X) \big)$.
For leaf-search, the output size and scaling factor are the same as in Repeated $\rho$-NN, and so the asymptotic complexity is $\mathcal{L} \coloneqq \mathcal{O} \left( k \cdot \bigg( 1 + 2 \cdot \Big( \frac{\hat{|C|}}{k} \Big) ^ {d^{-1}} \bigg)^d \right)$.

@@

We perform some simple auto-tuning to select the optimal $k$-NN algorithm to use with a given dataset.
We start by taking the center of every cluster at a low depth, (e.g. 10), in the cluster tree as a query.
This gives us a small, representative sample of the dataset.
Using these clusters' centers as queries, and a user-specified value of $k$, we record the time taken for $k$-NN search on the sample using each of the three algorithms described in Section~\ref{sec:methods:knn-search}.
We select the fastest algorithm over all the queries as the optimal algorithm for that dataset and value of $k$.
Note that even though we select the optimal algorithm based on use with some user-specified value of $k$, we still allow search with any value of $k$.

@@

To elaborate on the augmentation process, we start with an original dataset from the ANN-benchmarks suite.
Let $X$ be the dataset, $d$ be its dimensionality, $\epsilon$ be a user-specified noise level, and $m$ be a user-specified integer multiplier.
For each datum $x \in X$, we create $m - 1$ new data points within a distance $\epsilon \cdot ||x||$ of $x$ where $||x||$ is the Euclidean distance from $x$ to the origin.
We construct a random vector $r$ of $d$ dimensions in the hyper-sphere of radius $\epsilon$ centered at the origin.
We then add $r$ to $x$ to get a new point $x'$.
Since $||r|| \leq \epsilon$, we have that $||x - x'|| \leq \epsilon$ (i.e., $x'$ is within a distance $\epsilon$ of $x$).
This produces a new dataset $X'$ with $|X'| = m \cdot |X|$.
This augmentation process does not add to the overall topological structure of the dataset, but it does increase its cardinality by a factor of $m$.
This allows us to isolate the effect of cardinality on search performance from that of other factors such as dimensionality, choice of metric, or the topological structure of the dataset.

>>>>

We define a \textit{distance function} $f : \textbf{X} \times \textbf{X} \mapsto \mathbb{R}^+ \ \cup \ \{0\}$ which, given two points, deterministically returns a finite and non-negative real number.
A distance of zero defines identity among points (i.e.,\,$f(x, y) = 0 \Leftrightarrow x = y$) and larger values indicate greater dissimilarity among points.
We also require that the distance function be symmetric (i.e.,\,$f(x, y) = f(y, x) \ \forall \ x, y \in \textbf{X}$).
In addition to these constraints, if the distance function obeys the triangle inequality (i.e.,\,$f(x, y) \leq f(x, z) + f(z, y) \ \forall \ x, y, z \in \textbf{X}$), then it is also a \textit{distance metric}.
Similar to~\cite{yu2015entropy}, when used with distance metrics, all search algorithms in CAKES are exact.
For example, Euclidean, Levenshtein~\cite{levenshtein1966binary} and Dynamic Time Warping (DTW)~\cite{muller2007dynamic} distances are all distance metrics, while cosine distance is not a metric because it violates the triangle inequality (e.g.,\,consider the points $x = (1, 0)$, $y = (0, 1)$ and $z = (1, 1)$ on the Cartesian plane).

@@

Intuitively, LFD measures the rate of change in the number of points in a ball of radius $r$ around a point $q$ as $r$ increases. When the vast majority of points in the dataset have low ($\ll D$) LFD, we can simply say that the dataset has low LFD.
We stress that this concept differs from the \textit{embedding dimension} of a dataset.
To illustrate the difference, consider the SILVA 18S rRNA dataset which contains genomic sequences with unaligned lengths of up to 3,718 base pairs and aligned length of 50,000 base pairs.
Hence, the \textit{embedding dimension} of this dataset is at least 3,718 and at most 50,000.
However, physical constraints (namely, biological evolution and biochemistry) constrain the data to a lower-dimensional manifold within this space.
LFD is an approximation of the dimensionality of that lower-dimensional manifold in the ``vicinity'' of a given point.
Figure~\ref{fig:results:lfd-plots} illustrates this concept on a variety of datasets, showing how real datasets uphold the manifold hypothesis.
For real-world datasets, we expect the LFD to be locally uniform, i.e.,\,when $r$ is small, but potentially highly variable at larger length scales, i.e.,\,when $r$ is large.

@@

We define a \textit{cluster} as a set of points with a \textit{center} and a \textit{radius}.
The center is the geometric median of the points in the cluster, i.e.,\,it is the point which minimizes the sum of distances to all other points in the cluster.
In cases where the cardinality of the cluster is large, we take a random subsample of $\sqrt{|C|}$ points and compute the geometric median of that subsample~\cite{ishaq2019clustered}.
The center, therefore, is one of the points in the cluster and is used as a representative of the cluster.
The radius is the maximum distance from the center to any point in the cluster.
Each non-leaf cluster has two child clusters in much the same way that a node in a binary tree has two child nodes.
Note that clusters can have overlapping volumes and, in such cases, points in the overlapping volume are assigned to exactly one of the overlapping clusters.
As a consequence, a cluster can be a proper subset of the metric ball at the same center and radius, i.e.,\,$C(c, r) \subset B(c, r)$.

Hereafter, when we refer to the LFD of a cluster, it is estimated at the length scale of the cluster radius and half that radius, i.e.,\,using Equation~\ref{eq:methods:lfd-half}.
We also only use points that are in $C(c, r)$ instead of all those in $B(c, r)$.

@@

The \textit{radius} of $C$ is the maximum distance from the \textit{center} to any other point in $C$.
The point which is responsible for that radius (i.e.,\,the furthest point from the center) is designated the \textit{left pole} and the point which is furthest from the left pole is designated the \textit{right pole}.
We then partition the cluster into a \textit{left child} and a \textit{right child}, where the left child contains all points in the cluster which are closer to the left pole than to the right pole, and the right child contains all points in the cluster which are closer to the right pole than to the left pole.
Without loss of generality, we assign to the left child those points which are equidistant from the two poles.
Starting from a root cluster containing the entire dataset, we repeat this procedure until each leaf contains only one datum, or we meet some other user-specified stopping criteria, for example, minimum cluster radius, minimum cluster cardinality, maximum tree depth, etc.
This process is described in Algorithm \ref{alg:methods:partition}.
During the partitioning process, we also compute (and cache) the LFD of each cluster using Equation~\ref{eq:methods:lfd-half}.

@@

Note that Algorithm~\ref{alg:methods:partition} does not necessarily produce a balanced tree.
In fact, for real datasets, we expect anything but a balanced tree; the varying sampling density in different regions of the manifold and the low dimensional ``shape'' of the manifold itself will cause it to be unbalanced.
The only case in which we would expect a balanced tree is if the dataset were uniformly distributed, e.g.,\,in a $d$-dimensional hyper-cube.
In this sense, the imbalance in the tree is a feature, not a bug, as it reflects the underlying structure of the data.

@@

This complexity analysis ignores the cost of checking the continuation criteria.
While a user could specify criteria of any complexity (our implementation allows for this), for this paper we continue clustering until each cluster is a singleton, i.e.,\,it contains only one point or duplicates of the same point and has a radius of zero.
This criterion costs $\mathcal{O}(1)$ to check per cluster, and so does it not affect the overall complexity of the algorithm as used in this paper.

@@

Based on the assumption that the LFD near the query does not differ significantly from that of the clusters near the query (the set $Q$), Equation~\ref{eq:methods:repeated-rnn-factor} suggests that in the expected case, we need only two iterations of tree-search to find $k$ neighbors:
one iteration to find at least one cluster, and one more to find enough of the $k$ neighbors.
Since this is a constant factor, complexity of tree-search for Repeated $\rho$-NN is the same as that of $\rho$-NN search, i.e.,\,$\mathcal{O}\big(\log\mathcal{N}_{\hat{r}}(X)\big)$.

@@

In addition, by our assumption that the LFD at the query is not significantly different from the LFD of nearby clusters, we have that $\hat{d} \approx d$.
By combining the bounds for tree-search and leaf-search, we see that Repeated $\rho$-NN has an asymptotic complexity of
\begin{gather}
    \mathcal{O}
    \Bigg(
        \underbrace{
            \log~\overbrace{\mathcal{N}_{\hat{r}}(X)}^{\textrm{metric entropy}}
        }_{\textrm{tree-search}}
        \ + \
        \underbrace{
            \overbrace{k}^{\textrm{output size}}
            \overbrace{\bigg( 1 + 2 \cdot \Big( \frac{\hat{|C|}}{k} \Big) ^ {d^{-1}} \bigg)^d}^{\textrm{scaling factor}}
        }_{\textrm{leaf-search}}
    \Bigg)
    \label{eq:methods:repeated-rnn-complexity}
\end{gather}
where $\mathcal{N}_{\hat{r}}(X)$ is the metric entropy of the dataset, $d$ is the LFD of the dataset, and $k$ is the number of nearest neighbors.
We note that the scaling factor should be close to 1 unless fractal dimension is highly variable in the region around the query (i.e.,\,if $\hat{d}$ differs significantly from $d$).

@@

This process is described in Algorithm~\ref{alg:methods:depth-first-sieve}.
It terminates when $H$ is full and the top priority element in $H$ has $\delta$ less than the top priority element in $Q$, i.e.,\,the theoretically closest point left to be considered in $Q$ is farther from the query than the $k^{th}$ nearest neighbor in $H$.
This leaves $H$ containing exactly the $k$ nearest neighbors to the query.

@@

We start with the assumption that the LFD near the query does not differ significantly from that of nearby clusters, and we consider leaf clusters with cardinalities near $k$.
Let $d$ be the LFD in this region.
Then the number of leaf-clusters in $Q$ is bounded above by $2d$, e.g.,\,we could have a cluster overlapping the query ball at each end of each of $\lceil d \rceil$ mutually-orthogonal axes.
In the worst-case scenario for tree-search, these leaf clusters would all come from different branches of the tree, and so tree-search looks at $2 \cdot \lceil d \rceil \cdot \log \mathcal{N}_{\hat{r}}(X)$ clusters.
Thus, the asymptotic complexity is $\mathcal{T} \coloneqq \mathcal{O} \big( \lceil d \rceil \cdot \log \mathcal{N}_{\hat{r}}(X) \big)$.
For leaf-search, the output size and scaling factor are the same as in Repeated $\rho$-NN, and so the asymptotic complexity is $\mathcal{L} \coloneqq \mathcal{O} \left( k \cdot \bigg( 1 + 2 \cdot \Big( \frac{\hat{|C|}}{k} \Big) ^ {d^{-1}} \bigg)^d \right)$.

@@

We perform some simple auto-tuning to select the optimal $k$-NN algorithm to use with a given dataset.
We start by taking the center of every cluster at a low depth, (e.g.,\,10), in the cluster tree as a query.
This gives us a small, representative sample of the dataset.
Using these clusters' centers as queries, and a user-specified value of $k$, we record the time taken for $k$-NN search on the sample using each of the three algorithms described in Section~\ref{sec:methods:knn-search}.
We select the fastest algorithm over all the queries as the optimal algorithm for that dataset and value of $k$.
Note that even though we select the optimal algorithm based on use with some user-specified value of $k$, we still allow search with any value of $k$.

@@

To elaborate on the augmentation process, we start with an original dataset from the ANN-benchmarks suite.
Let $X$ be the dataset, $d$ be its dimensionality, $\epsilon$ be a user-specified noise level, and $m$ be a user-specified integer multiplier.
For each datum $x \in X$, we create $m - 1$ new data points within a distance $\epsilon \cdot ||x||$ of $x$ where $||x||$ is the Euclidean distance from $x$ to the origin.
We construct a random vector $r$ of $d$ dimensions in the hyper-sphere of radius $\epsilon$ centered at the origin.
We then add $r$ to $x$ to get a new point $x'$.
Since $||r|| \leq \epsilon$, we have that $||x - x'|| \leq \epsilon$ (i.e.,\,$x'$ is within a distance $\epsilon$ of $x$).
This produces a new dataset $X'$ with $|X'| = m \cdot |X|$.
This augmentation process does not add to the overall topological structure of the dataset, but it does increase its cardinality by a factor of $m$.
This allows us to isolate the effect of cardinality on search performance from that of other factors such as dimensionality, choice of metric, or the topological structure of the dataset.

### 2025-01-05 sections/3-datasets.tex
wip: changes by NI
dealt with e.g., i.e., dashes

\section{Datasets And Benchmarking}
\label{sec:datasets-and-benchmarks}

@@

To demonstrate CAKES with a more exotic distance function, we also benchmarked on the SILVA 18S ribosomal RNA dataset~\cite{10.1093/nar/gks1219}.
This dataset contains ribosomal RNA sequences of 2,224,640 genomes with an aligned length of 50,000 letters.
We held out a set of 1,000 random sequences from the dataset to use as queries for benchmarking.
We use Levenshtein~\cite{levenshtein1966binary} distance on the unaligned sequences to build the tree and to perform $k$-NN search.
Since this dataset contains many redundant, i.e. nearly identical, sequences, we used a maximum tree-depth of 128 as a partition criterion in Algorithm~\ref{alg:methods:partition}.

>>>>

\section{Datasets And Benchmarks}
\label{sec:datasets-and-benchmarks}

@@

To demonstrate CAKES with a more exotic distance function, we also benchmarked on the SILVA 18S ribosomal RNA dataset~\cite{10.1093/nar/gks1219}.
This dataset contains ribosomal RNA sequences of 2,224,640 genomes with an aligned length of 50,000 letters.
We held out a set of 1,000 random sequences from the dataset to use as queries for benchmarking.
We use Levenshtein~\cite{levenshtein1966binary} distance on the unaligned sequences to build the tree and to perform $k$-NN search.
Since this dataset contains many redundant, i.e.,\,nearly identical, sequences, we used a maximum tree-depth of 128 as a partition criterion in Algorithm~\ref{alg:methods:partition}.
